---
title: "STOR 320 Tutorial on Programming 1"
author: ""
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=F)
options(scipen=999)
library(tidyverse)    #Essential Functions
library(Ecdat)        #Contains Economic Datasets
```

# Introduction

In this tutorial, we will try to understand some fundamental control structures used in statistical programming. In the beginning, we will separately analyze different control structures. 

# Part 1: *If-Else* Statements
General Construction:

- `if`:
```{r, eval=F}
if (CONDITION) {
	ACTION
}
```

- `if else`:
```{r, eval=F}
if (CONDITION) {
	ACTION 1
} else {
	ACTION 2
}
```

- `ifelse()`:

```{r, eval=F}
ifelse(CONDITION,ACTION1,ACTION2)
```

## Chunk 1: Illustration of *If*
```{r,eval=F}
x = 3
if(x > 0){
  print(log(x))
}

x = -3
if(x > 0){
  print(log(x))
}
```


## Chunk 2: Illustration of *If-Else*
```{r,eval=F}
x = 3
if(x > 0){
  print(log(x))
} else{
  message("Unable to Take Logarithm")
}

x = -3
if(x > 0){
  print(log(x))
} else {
  message("Unable to Take Logarithm")
}
```

### Excercise 1
Write code that takes numbers `a` and `b` as input and prints 'a is greater than b' if `a>b`, otherwise prints 'a is not greater than b'.

```{r}
a = 10
b = 8
# write code here
```

## Chunk 3: Potential Problem of *If-Else* Statements
```{r,eval=F}
x = BLANK
if(x > 0){
  print(log(x))
}

if(x > 0){
  print(log(x))
} else{
  message("Unable to Take Logarithm")
}
```


## Chunk 4: Fixing Potential Problem in Chunk 3
```{r,eval=F}
x=BLANK
if(CONDITION){
  if(x > 0){
    print(log(x))
  } else{
    message("Unable to Take Logarithm")
  }
} else{
  message("Please Input Numbers")
}
```

### Excercise 2
Redo Excercise 1 but check the data types before doing the comparison. Hint: `&&` (and) and `||` (or) can be used to combine multiple logical expressions. Please don't use `&` and `|` in an `if` statement: these are vectorized operations.
```{r}
a = '10'
b = 8
#
```

## Chunk 5: Vectorized Version with `ifelse()`
```{r,eval=F}
x=c(-1,3,200)
print(log(x))

y1 =  if(x > 0){
        log(x)
      } else{
        NA
      }
print(y1)
```

```{r, eval=F}
y2 = ifelse(x>0,log(x),NA)
print(y2)
```

## Chunk 6: Nested `ifelse()` Statements
```{r,eval=F}
x=rnorm(1000,mean=0,sd=1)
y=ifelse(abs(x)<1,"Within 1 SD",ifelse(abs(x)>2,"Far Far Away","Between 1 and 2 SD"))
y.fct=factor(y,levels=c("Within 1 SD","Between 1 and 2 SD","Far Far Away"))
ggplot() +
  geom_bar(aes(x=y.fct),fill="lightskyblue1") +
  theme_minimal()
```

### Excercise 3
Please use `ifelse()` function to create a new column `WageLevel` with `lwage` column of the `Wages` dataset from `Ecdat`. 

- If the `lwage` is greater than $7.0$, the corresponding value in the final vector is `High Wage`. 
- If `lwage` is lower than $6.4$, the value is `Low Wage`. 
- For all the other cases, the values are `Normal`.

```{r}
#
```

# Part 2: Loops

## Chunk 1: Checking Geometric Series Proof with *for* loop
Geometric Series:
$a, ar, ar^2, ar^3,...$

Formula of Sum:
$\sum_{k=0}^\infty ar^k=\frac{a}{1-r}$, for $|r|<1$.

```{r,eval = F}
a=BLANK #Any Number
r=BLANK #Any Number Between -1 and 1: abs(r)<1

theoretical.limit=a/(1-r)

START=a

FINISH.1 = START + a*r^1

FINISH.2 = FINISH.1 + a*r^2

FINISH.3 = FINISH.2 + a*r^3

FINISH.10 = a 
for(k in 1:10){
  FINISH.10=FINISH.10+a*r^k
}

FINISH.100 = a 
for(k in 1:100){
  FINISH.100=FINISH.100+a*r^k
}

DATA = tibble(k=c(1,2,3,10,100,"Infinity"),
            SUMMATION=c(FINISH.1,FINISH.2,FINISH.3,
                        FINISH.10,FINISH.100,
                        theoretical.limit))
print(DATA)

ABSOLUTE.ERROR = abs(FINISH.100-theoretical.limit)
print(ABSOLUTE.ERROR)
```

### Excercise 4
```{r}
set.seed(4)
u = rnorm(100)
```

Use `for` loop to calculate sum of the squares of the first 10 elements of vector `u`. Hint: `^` (Exponents), `3^2=9`
```{r}
#
```

Use `for` loop and `if` statement to calculate the sum of squares of the elements with even indices of vector `u`. Hint: `%%` gives the remainder of division.
```{r}
#
```

## Chunk 2: Checking Geometric Series Proof with *while* loop
```{r,eval=F}
a=BLANK
r=BLANK

FINISH=a
k=0
while(abs(FINISH-a/(1-r)) > 1e-10) {
  k=k+1
  FINISH = FINISH + a*r^k
  #if(k>100) break
}
print(c(k,FINISH))
```

## Chunk 3: Saving Steps in Geometric Series for Figure
```{r,eval=F}
a=10
r=-0.75
theoretical.limit=a/(1-r)

K=10 #How Many Steps Do You Want to Save?

summation=rep(NA,(K+1))
summation[1]=a
for (k in 1:K) {
  summation[k+1]=summation[k] + a*r^k
}

ggplot() +
  geom_line(aes(x=1:(K+1),y=summation)) +
  geom_hline(yintercept=theoretical.limit,
             linetype="dashed")
```

### Excercise 5
Write `for` loops to generate 100 random samples from normal distributions with means of 0 to 99 and save the random samples to a vector `a`.

- The `k`th component of `a` is generated from $N(k-1,1)$.
- Hint: Function for generating random samples from normal distribution `rnorm`.

```{r}
set.seed(100)
#
```