---
title: "STOR 320 Tutorial on Programming 2"
author: ""
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=F)
options(scipen=999)
library(tidyverse)    #Essential Functions
library(Ecdat)        #Contains Economic Datasets
```

# Introduction

In this tutorial, we will try to understand some fundamental control structures used in statistical programming. In the beginning, we will separately analyze different control structures. 

# Part 1: Loops

## Chunk 1: Correlation Matrix for Variables in `Cigar` from `Ecdat`
```{r,eval=F}
head(Cigar)
print(round(cor(Cigar),3))
```

## Chunk 2: Using a Double Loop to Create a Correlation Matrix
```{r,eval=F}
CORR.CIGAR1=matrix(NA,9,9) #Empty Matrix Initialized
rownames(CORR.CIGAR1)=names(Cigar)
colnames(CORR.CIGAR1)=names(Cigar)

seq_along(names(Cigar))

for(j in seq_along(names(Cigar))){
  for(k in seq_along(names(Cigar))){
    CORR.CIGAR1[j,k]=cor(Cigar[,j],Cigar[,k])
  }
}
print(round(CORR.CIGAR1,3))
```

```{r, eval=F}
CORR.CIGAR2=matrix(NA,8,8)
rownames(CORR.CIGAR2)=names(Cigar)[-1]
colnames(CORR.CIGAR2)=names(Cigar)[-1]
for(j in 1:BLANK){
  for(k in 1:BLANK){
    CORR.CIGAR2[BLANK,BLANK]=cor(Cigar[,BLANK],Cigar[,BLANK])
  }
}
print(round(CORR.CIGAR2,3))
```

## Chunk 3: Correlation Matrix for Variables in `HI` from `Ecdat`
```{r,eval=F}
HealthInsurance=HI
head(HealthInsurance)
#?HI
#print(round(cor(HealthInsurance),3)) #Try this Code
```

## Chunk 4: Using a Double Loop to Compute *5*-Number Summary for Numeric Variables
```{r,eval=F}
var.names = names(HealthInsurance)

FiveSum.HI = matrix(NA,length(var.names),6)
colnames(FiveSum.HI) = c("Variable","Min","Q1","Q2","Q3","Max")

for(VAR in seq_along(var.names)){
  if(is.numeric(HealthInsurance[,VAR])){
    MIN=min(HealthInsurance[,VAR])
    Q1=quantile(HealthInsurance[,VAR],0.25)
    Q2=median(HealthInsurance[,VAR],0.5)
    Q3=quantile(HealthInsurance[,VAR],0.75)
    MAX=max(HealthInsurance[,VAR])
    FiveSum.HI[VAR,]=c(names(HealthInsurance)[VAR],MIN,Q1,Q2,Q3,MAX)
  } else {
    cat("Variable",var.names[VAR],"is not numeric\n")
    FiveSum.HI[VAR,]=c(names(HealthInsurance)[VAR],rep(NA,5))
  }
}
print(as.tibble(na.omit(FiveSum.HI)))
```


```{r, eval=F}
FiveSum.HI2 = NULL
Numeric.names = NULL
for(VAR in seq_along(var.names)){
  if(is.numeric(HealthInsurance[,VAR])){
    MIN=min(HealthInsurance[,VAR])
    Q1=quantile(HealthInsurance[,VAR],0.25)
    Q2=median(HealthInsurance[,VAR],0.5)
    Q3=quantile(HealthInsurance[,VAR],0.75)
    MAX=max(HealthInsurance[,VAR])
    FiveSum.HI2=rbind(FiveSum.HI2,c(MIN,Q1,Q2,Q3,MAX))
    Numeric.names=c(Numeric.names,var.names[VAR])
  } 
}
FiveSum.HI3=as.tibble(cbind(Numeric.names,as.tibble(FiveSum.HI2)))
names(FiveSum.HI3) = c("Variable","Min","Q1","Q2","Q3","Max")
print(FiveSum.HI3)
```

### Excercise 1
The following code creates a new data frame `HI_Num` that only contains numeric columns in `HealthInsurance`.
```{r, eval=F}
HI_Num = HealthInsurance %>%
  select(c(1,8,9,10,11,13))
```

Using a Double Loop to Create a Correlation Matrix for numeric variables in `HealthInsurance`.
```{r}
#
```

### Excercise 2
Write a loop that loops over the columns of data `Wages` and reports the mean of the column if it is numeric and the total number of unique characters if it’s a character vector.
```{r}
#
```

### Excercise 3
Writing a function to calculate five summary statistics and generating five summary statistics for numeric variables in `HealthInsurance`.

```{r}
#
```

# Part 2: Simple Random Sampling

## Chunk 1: Sampling from Known Distributions
```{r,eval=F}
x1=rnorm(100,mean=82,sd=2)
ggplot()+geom_histogram(aes(x1))

x2=rbinom(100,size=10,prob=1/6)
ggplot()+geom_bar(aes(x2))
```

## Chunk 2: Experiment for Flipping Coins 
```{r,eval=F}
prop=rep(NA,1000)
for(k in 1:1000){
  set.seed(k)
  x=sample(c("H","T"),size=k,replace=T,prob=c(BLANK,1-BLANK))
  prop[k]=mean(x=="H")
}
ggplot() + 
  geom_line(aes(x=1:1000,y=prop),alpha=0.5) + 
  geom_hline(yintercept=BLANK,linetype="dashed",color="red",size=2) +
  theme_minimal()
```



